Score-based tests of measurement invariance: Use in practice

### 2013-08-21 frontiers/frontiers.Rnw
Added to frontiers paper: o Added detail to background o Revised tutorial discussion o Revised code in tutorial
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@240 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Tests of measurement invariance via stochastic processes:
  {S}ome properties}
\twoauthors{Ting Wang and Edgar C.\ Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being straightforward.
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters appear to vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      l(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $l(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
of these scores forms the ``meat'' for the calculation of robust
(Huber-White) standard errors. %TODO cite

To expand on the above equation, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  Scores close to zero
indicate a ``good'' description, and scores far from zero indicate a
``bad'' description.  Each of an individual's $k$ scores represent one
model parameter, so that the scores provide specific information about
each parameter's fit to each individual.

To use these scores for testing, we order individuals according to
an auxiliary variable (the variable against which we are testing 
measurement invariance) and look for trends in the
scores (roughly speaking).  For example, imagine that we are testing
for measurement 
invariance w.r.t.\ age.  If the manifest variables violate measurement
invariance here, then some parameter estimates may be too large for
young individuals and too small for old individuals (say).  This
result would be reflected in the scores, where young individuals'
scores would be greater than zero and old individuals' scores would be
less than zero (though the sign of the scores will also depend on whether a
function is being minimized or maximized).  Conversely, if measurement
invariance holds, then all 
individuals' scores will fluctuate around zero.

To formalize the ideas in the previous paragraph, we focus on a cumulative
sum of the ordered scores.  This cumulative sum may be written as
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ (i.e., a floor
operator) and $x_{(i)}$ reflects the individual with the
$i^{\text{th}}$-smallest value of the auxiliary variable.
We specifically focus on how the
cumulative sum fluctuates as more individuals' scores are added to it,
starting with the youngest and ending with the oldest individual
(assuming that age is the auxiliary variable of interest).  
The summation is also typically premultiplied by the inverse square
root of the information matrix, which serves to decorrelate the
fluctuation processes associated with individual model parameters.

Under the
hypothesis of measurement invariance, a central limit theorem can be
used to show that the fluctuation of the above cumulative sum follows a
Brownian bridge.  This result allows us to calculate $p$-values and
critical values for test statistics under the hypothesis of
measurement invariance.  We can obtain test statistics associated with
all model parameters, as well as test statistics associated with
subsets of model parameters.

%% TODO focus on ordinal
Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores.  One could
take the absolute maximum that the cumulative sum attains for any
parameter of interest, resulting in a {\em double max} statistic.
Alternatively, one could sum the (squared) cumulative sum across parameters
of interest and take the maximum across individuals, resulting in a
{\em maximum Lagrange multiplier} statistic 
\cite<see>[for further discussion of this idea]{MerZei13}.
%% TODO equations of statistics?

Importantly, these statistics were derived for situations where
individuals are uniquely ordered according to the auxiliary variable.
This is not always the case for measurement invariance applications,
where the auxiliary variable is often ordinal.  To remedy this
situation, \citeA{MerFanZei} extended the framework to situations
where one has an ordinal auxiliary variable of interest.  Essentially,
one allows all individuals with the same value of the auxiliary
variable to enter into the cumulative sum at the same time.  Test
statistics analogous to those describe above are then computed, with
modified critical values being adopted to reflect the change in the
statistics' computation.

\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
\section{Tutorial}
In general, a set of scales is defined to be measurement invariant with respect
to an auxiliary variable $V$ if:

\begin{equation}
\label{equ:invariance}
f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
\end{equation}

where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
for variable $i$ that the scales purport to measure, and $f$ is the model's 
distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
of measurement invariance is

\begin{equation}
\label{equ:null}
H_{0}: \theta_{A}=\theta_{B},
\end{equation}

The alternative hypothesis is

\begin{equation}
\label{equ:alternative}
H_{1}:\theta_{i}=
\begin{cases}
\theta^{(A)} & v_{i}\leq v\\
\theta^{(B)} & v_{i}\geq v
\end{cases}
\end{equation}

The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
(group A and group B) can be obtained by Maximum Likelihood(ML) 
from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
respectively. The LR test statistic for the given threshold $v$ is then

\begin{equation}
\label{equ:LRT}
LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
{(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
}],
\end{equation}

,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
to the number difference of parameters in $\theta$.

To be more specific, in the following section, we will illustrate with an example to demonstrate how to implement multi group analysis in lavaan package. For more details, see \cite{vande12}

\subsection{Multiple Group Analysis}
Froh et al\citeA{FroFan11} had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
including Gratitude Questionnaire 6(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective Checklist(GAC; \citeNP{MccEmm02})  and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); \citeNP{ThoWat03}). In this specific analysis, the authors were interested in whether the youth factor 
structure for GQ-6 is a tau equivalent model (factor loadings restricted to be equal across
levels of age)or a congeneric model(all parameters free for each level of age).

To conduct the analysis, first of all, we need to read in data to R and clean the data.
<<read data>>=
# Reading in the data
data("YouthGratitude",package="psychotools")
# remove cases with 'imputed' values(not in 1,...,9)
yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
@
Second, we want to test which parameter(s) violate measurement invariance. 
In lavaan package, loading, manifest variables' intercept and factor means 
could be tested together or specifically. For more details, 
please see \cite{vande12}. Since in most cases researchers are 
interested in the factor loading invariance, we will use that as the 
example in this tutorial.
To implement LRT as ~\ref{equ:LRT}, we need to get estimation 
for $\hat{\theta}$, without differing groups.

<<theta estimate>>=
# estimate theta 
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
anova(theta, thetaAB)
@
Likelihood test is sensitive to large observations (CITE), it is not surprising to 
have a significant result, although Froh et al\citeA{FroFan11} showed that tau equivalent
model has as good alternative fit indices as congeneric model.

\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large. However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure described in
the Background section.  That is, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
values of $V$.

%% TO BE MOVED TO BACKGROUND
%% In practice, there are three test statistics can be used: 
%% double maximum, CvM and maxLM. 
%% \begin{equation}
%% DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
%% \end{equation}
%% \begin{equation}
%% CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
%% \end{equation}
%% \begin{equation}
%% maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
%% \end{equation}

To demonstrate these tests in R, we will use the youth gratitude data
and assume that age group is continuous.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to 
get the $B(\hat{\theta}_{ij})$. After that, we can get the 
three test statistics' significance by using sctest function.

<<Get Estimate>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@
In applying the proposed continuous test, the $p$ value for 
CvM and maxBB $p$ value is larger than likelihood test. 
Besides, CvM $p$ value is not significant.

\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the statistics value,
corresponding $p$ value and the plot ordered by age level.

\begin{figure}
\caption{Get Continuous Estimate DM Alternative Way}
\label{fig:conti1res}
\setkeys{Gin}{width=\textwidth}
<<conti1res, fig=TRUE, height=7, width=7>>=
DM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
DM
@
\end{figure}

\begin{figure}
\caption{Get Continuous Estimate CvM Alternative Way}
\label{fig:conti2res}
\setkeys{Gin}{width=\textwidth}
<<conti2res, fig=TRUE, height=7, width=7>>=
CvM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
CvM
@
\end{figure}

<<conti3resmaxLM>>=
maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@
Since there are more than one observations located in each age level. 
So the observation order within each age level is not unique, thus, 
maxLM can't produce the plot. Actually, all three continuous statistics 
were calculated on the arbitrary order that particular dataset entered 
in each age level. In order to accurately account for the multiple observations 
at the same age level, it is more appropriate to use the proposed ordinal test.

\subsection{Ordinal Test}
As a matter of fact, there is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable. That is, instead 
of aggregating over all $i=1,...n$ individuals, we 
first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
associated with the first $m-1$ levels of $V$. 
%% TODO Move these details to Background section
We then aggregate the cumulative scores only over 
$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal 
test is that we don't force arbitrary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more real status of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

@@

Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
@
\end{figure}

\begin{figure}
\caption{Get Ordinal Estimate maxLMo Alternative Way}
\label{fig:ord2res}
\setkeys{Gin}{width=\textwidth}
<<ord2res, fig=TRUE, height=7, width=7>>=
maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
               fnl=opval1, fplot=TRUE)
maxLMo 
@
\end{figure}

>>>>

\title{Score-based tests of measurement invariance: Use in practice}
\twoauthors{Ting Wang and Edgar C.\ Merkle}{Achim Zeileis}
\twoaffiliations{University of Missouri}{Universit\"{a}t Innsbruck}

@@

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being conceptually straightforward. 
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

Following model estimation, the tests primarily work on the {\em
  scores} of the fitted model; these are defined as
%% TODO indicate that this is evaluated at theta hat
\begin{equation}
\label{eq:score}
s(\hat{\theta};x_{i}) = \left ( \frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{1})},...,\frac{\partial
      \ell(\theta;x_{i})}{\partial(\theta_{k})} \right )^{T},
\end{equation} 
where $\ell(\theta; x_i)$ is the likelihood associated with individual
$i$, $\theta$ is a $k$-dimensional parameter vector, and
$\hat{\theta}$ is the maximum likelihood estimate.  The cross-product
of these scores forms the ``meat'' for the calculation of robust
(Huber-White) standard errors. %TODO cite

To verbally describe the above equation, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  Scores close to zero
indicate a ``good'' description, and scores far from zero indicate a
``bad'' description.  Each of an individual's $k$ scores represent one
model parameter, so that the scores provide specific information about
each parameter's fit to each individual.

To use these scores for testing, we order individuals according to
an auxiliary variable (the variable against which we are testing 
measurement invariance) and look for ``trends'' in the
scores.  For example, imagine that we are testing
for measurement 
invariance w.r.t.\ age.  If the manifest variables violate measurement
invariance here, then some parameter estimates may be too large for
young individuals and too small for old individuals (say).  This
result would be reflected in the scores, where young individuals'
scores may be greater than zero and old individuals' scores 
less than zero (though the sign of the scores will depend on whether a
function is being minimized or maximized).  Conversely, if measurement
invariance holds, then all 
individuals' scores will fluctuate around zero.

To formalize the ideas in the previous paragraph, we define a cumulative
sum of the ordered scores.  This cumulative sum may be written as
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ (i.e., a floor
operator) and $x_{(i)}$ reflects the individual with the
$i^{\text{th}}$-smallest value of the auxiliary variable.
We specifically focus on how the
cumulative sum fluctuates as more individuals' scores are added to it,
starting with the youngest and ending with the oldest individual
(assuming that age is the auxiliary variable of interest).  
The summation is also typically premultiplied by the inverse square
root of the information matrix, which serves to decorrelate the
fluctuation processes associated with individual model parameters (i.e., the process associated with one parameter is not correlated with the processes associated with other parameters).

Under the
hypothesis of measurement invariance, a central limit theorem can be
used to show that the fluctuation of the above cumulative sum follows a
Brownian bridge.  This result allows us to calculate $p$-values and
critical values for test statistics under the hypothesis of
measurement invariance.  We can obtain test statistics associated with
all model parameters and with subsets of model parameters.

%% TODO focus on ordinal
Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores.  For example, one could
take the absolute maximum that the cumulative sum attains for any
parameter of interest, resulting in a {\em double max} statistic (the maximum is taken across parameters and individuals).
Alternatively, one could sum the (squared) cumulative sum across parameters
of interest and take the maximum or the average across individuals, resulting in a {\em maximum Lagrange multiplier} statistic and Cram\'{e}r-von Mises statistic, respectively \cite<see>[for further discussion]{MerZei13}.  These statistics are given by
\begin{eqnarray}
    \label{eq:dmax}
    \mathit{DM} & = & \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} \\
        \label{eq:cvm}
    \mathit{CvM}     & = & n^{-1} \sum_{i = 1,\dots, n} \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2, \\
    \label{eq:maxlm}
    \max \mathit{LM} & = & \max_{i = \underline{i}, \dots, \overline{\imath}} ~
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
      \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2.
\end{eqnarray}

Critical values associated with $\mathit{DM}$ can be obtained analytically, while critical values associated with the other statistics can be obtained from direct simulation \cite{Zei06} or from
more refined techniques \cite{han97}.  This issue should not be important to the user, as critical values are obtained directly from the R implementation described later.

Importantly, the above statistics were derived for situations where
individuals are uniquely ordered according to the auxiliary variable.
This is not always the case for measurement invariance applications,
where the auxiliary variable is often ordinal.  To remedy this
situation, \citeA{MerFanZei} extended the framework to situations
where one has an ordinal auxiliary variable of interest.  Essentially,
one allows all individuals with the same value of the auxiliary
variable to enter into the cumulative sum at the same time.  Analogous test 
statistics are then computed, with
modified critical values being adopted to reflect the change in the
statistics' computation.

For an ordinal auxiliary variable with $m$ levels, these modifications are based on $t_{\ell}$ $(\ell=1,\ldots,m-1)$, which are the empirical, cumulative proportions of individuals observed at the first $m-1$ levels.
The modified statistics are then given by
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}      
                             \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~
                             \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}
                             \sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}
where $i_{\ell}=\lfloor n \cdot t_{\ell} \rfloor$ $(\ell=1,\ldots,m-1)$.
Critical values associated with the $\mathit{WDM}_o$ statistic can be obtained directly from a multivariate normal distribution \cite<see>{hotzei08}, while critical values associated with $\max \mathit{LM}_o$ can be obtained via simulation.  This simulation is somewhat computationally intensive and, in practice, takes about 10 minutes on the authors' computers.  However, the wait appears to be worth it, as \citeA{MerFanZei} found the performance of the $\max \mathit{LM}_o$ statistic to be considerably better than the $\mathit{WDM}_o$ statistic.

\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
\section{Tutorial}

In this section, we demonstrate how the above tests can be carried out in~R, using the package lavaan \cite{lavaan11} for model estimation and strucchange \cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning the applicability of adult gratitude scales to youth, available in the~R package psychotools CITE.  The data consist of responses to three adult gratitude scales from $n=1401$ youth aged 10--19 years. The original authors were specifically interested in whether the scales were measurement invariant across age.  Because the sample size at each age was unbalanced, the authors created age groups of approximately equal sample size.  In the examples below, we test for measurement invariance across these age groups.

We focus on measurement invariance of the factor loadings associated with the GQ-6 scale, using a one-factor model.
While the ``age group'' variable against which we are testing measurement invariance is best considered ordinal, for demonstration we consider its treatment as categorical, continuous, and ordinal.  Each of these treatments is described below in a separate section.

%% THE FOLLOWING DETAILS ARE REMOVED FOR NOW
%% In general, a set of scales is defined to be measurement invariant with respect
%% to an auxiliary variable $V$ if:

%% \begin{equation}
%% \label{equ:invariance}
%% f(x_{i}|t_{i},v_{i},...)=f(x_{i}|t_{i},...),
%% \end{equation}

%% where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
%% for variable $i$ that the scales purport to measure, and $f$ is the model's 
%% distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
%% gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
%% are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
%% of measurement invariance is

%% \begin{equation}
%% \label{equ:null}
%% H_{0}: \theta_{A}=\theta_{B},
%% \end{equation}

%% The alternative hypothesis is

%% \begin{equation}
%% \label{equ:alternative}
%% H_{1}:\theta_{i}=
%% \begin{cases}
%% \theta^{(A)} & v_{i}\leq v\\
%% \theta^{(B)} & v_{i}\geq v
%% \end{cases}
%% \end{equation}

%% The parameter estimate $\hat{\theta}^{(A)}$ , $\hat{\theta}^{(B)}$ ,$\hat{\theta}$ 
%% (group A and group B) can be obtained by Maximum Likelihood(ML) 
%% from observations $i=1,...m(v_{i}\leq v),i=m+1,...n(v_{i}\geq v), i=1,...,n$,
%% respectively. The LR test statistic for the given threshold $v$ is then

%% \begin{equation}
%% \label{equ:LRT}
%% LR(v)=-2[\ell(\hat{\theta};x_{1},,,,x_{n})-{\ell(\hat{\theta}^
%% {(A)};x_{1},...,x_{m})+\ell(\hat{\theta}^{(B)};x_{m+1},...,x_{n})
%% }],
%% \end{equation}

%% ,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
%% to the number difference of parameters in $\theta$.

\subsection{Categorical Treatment}

\fixme{Need to update the citation below so that the last name is ``van de Schoot''.}
Measurement invariance is most often tested using multiple groups models \cite<see, e.g., >{vande12}.  This amounts to assuming that our auxiliary variable is categorical (i.e., unordered), which is not true for the age groups in the data.  However, we demonstrate the procedure for completeness.

To conduct the analysis, we first load the data and keep only complete cases for simplicity (though the tests can be applied to incomplete data).
<<data>>=
## Reading in the data
data("YouthGratitude",package="psychotools")

## remove cases with 'imputed' values(not in 1,...,9)
compcases <- apply(YouthGratitude[,4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases,]
@
Next, we fit two models in lavaan: a one-factor model where loadings are restricted to be equal across age groups, and a one-factor model where loadings are free across age groups.
<<modest>>=
## Restricted model
reduced <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
             data=yg, group="agegroup", meanstructure=TRUE,
             group.equal="loadings")

## Unrestricted model
full <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
             data=yg, group="agegroup", meanstructure=TRUE)
@
Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ function, which implies that the GQ-6 violates measurement invariance.
<<LRT>>=
## LRT
anova(full, reduced)
@
Because our sample size is large, this likelihood ratio test is sensitive to small measurement invariance violations \cite{Ben80}.  Thus, the above test may be significant solely due to the large sample size.  Additionally, the above test is sensitive to small measurement invariance violations that are not likely to be of interest to researchers.  For example, imagine that the 15-year-olds' parameters are slightly different than the other age groups.  The 15-year-olds are in the middle of the age groups, and there is not likely to be any theoretical justification for 15-year-olds differing from every other age group.  We need test statistics that are sensitive to measurement invariance violations that are monotonic with age.  The score-based statistics described previously possess this property.

%% TODO Illustrate catL2BB statistic?

\subsection{Continuous Treatment}
If we are interested in measurement invariance violations that are monotonic with the age groups, it is perhaps simplest to treat the age groups as continuous.  In doing so, we can use the statistics from Equations~\eqref{eq:dmax},~\eqref{eq:cvm}, and~\eqref{eq:maxlm}.  That is, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
age (where age ties are broken arbitrarily).
This is demonstrated below, with similar code being useful when one is testing measurement invariance w.r.t.\ other continuous variables.

%% The multiple group test performs typically well when the $V$ 
%% is a relatively-small number of categories and the sample size 
%% is not too large. However, when $V$ (in this case, age) continuous variable, however, 
%% multiple-group analysis usually cannot be performed because there are 
%% no existing groups. Instead, we can follow the procedure described in
%% the Background section.

To proceed, we focus only on the restricted model that was fit in the previous section (the object \verb+reduced+).  We obtain scores associated with the factor loadings from this model, then test the fluctuation of these scores across \verb+agegroup+.  As demonstrated below, the score computation and testing can be efficiently carried out using \verb+sctest.lavaan()+.
%% Ting: I removed this because the model was already fit in the
%%       previous section
%%<<theta estimate>>=
%%theta<-cfa(
%%  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
%%  data=yg,group="agegroup",meanstructure=TRUE,
%%  group.equal="loadings")
%%@

<<Get Estimate>>=
dm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM")
cvm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM")
maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")

c(dm$p.value, cvm$p.value, maxlm$p.value)
@
In the code above, the \verb+parm=1:4+ argument contains parameter numbers reflecting the factor loadings that we wish to test.  The ordering of parameters can be obtained via the command \verb+coef(reduced)+.  Two of the three $p$-values output at the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

%% Aug 21: Ed stopped here.  TODO: graphs

\begin{figure}
\caption{Get Continuous Estimate DM Alternative Way}
\label{fig:conti1res}
\setkeys{Gin}{width=\textwidth}
<<conti1res, fig=TRUE, height=7, width=7>>=
DM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
DM
@
\end{figure}

\begin{figure}
\caption{Get Continuous Estimate CvM Alternative Way}
\label{fig:conti2res}
\setkeys{Gin}{width=\textwidth}
<<conti2res, fig=TRUE, height=7, width=7>>=
CvM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
CvM
@
\end{figure}

<<conti3resmaxLM>>=
maxLM<-sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@
Since there are more than one observations located in each age level. 
So the observation order within each age level is not unique, thus, 
maxLM can't produce the plot. Actually, all three continuous statistics 
were calculated on the arbitrary order that particular dataset entered 
in each age level. In order to accurately account for the multiple observations 
at the same age level, it is more appropriate to use the proposed ordinal test.

\subsection{Ordinal Test}
As a matter of fact, there is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable.

The difference between the continuous test and ordinal 
test is that we don't force arbitrary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more real status of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(reduced,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

@@

Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7>>=
WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
@
\end{figure}

\begin{figure}
\caption{Get Ordinal Estimate maxLMo Alternative Way}
\label{fig:ord2res}
\setkeys{Gin}{width=\textwidth}
<<ord2res, fig=TRUE, height=7, width=7>>=
maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
               fnl=opval1, fplot=TRUE)
maxLMo 
@
\end{figure}
